---
title: "Passing Efficiency and Team Success: A Simple Linear Regression Report of NFL Team Statistics"
author: "By Prasanna Sai Rohit Durbha"
abstract: "Modern American football analytics stress on efficiency rather than raw totals. In particular, the metric “expected points added” (EPA) summarises how much each play changes a team’s scoring expectation, making it a crucial way to calculate a team's offensive performance. This study asks whether teams that are more efficient on passing plays tend to out‑score their opponents over an NFL season. Using the publicly available National Football League team statistics dataset, which contains season‑level outcomes and advanced metrics such as average offensive EPA per pass play [@score_nfl], we fit a simple linear regression model with score differential (points scored minus points allowed) as the response and average offensive EPA per pass as the predictor. The fitted model shows a strong positive association: a 0.1‑point increase in EPA per pass predicts roughly a 57‑point improvement in season‑long score differential. Although passing efficiency explains about half of the variability in point margins, substantial residual variation remains, highlighting the importance of other factors such as defence and special teams."
format:
  pdf:
    number-sections: true
bibliography: references.bib
citeproc: true
---

## Abstract

Modern American football analytics stress on efficiency rather than raw totals. In particular, the metric “expected points added” (EPA) summarises how much each play changes a team’s scoring expectation, making it a crucial way to calculate a team's offensive performance. This study asks whether teams that are more efficient on passing plays tend to out‑score their opponents over an NFL season. Using the publicly available National Football League team statistics dataset, which contains season‑level outcomes and advanced metrics such as average offensive EPA per pass play [@score_nfl], we fit a simple linear regression model with score differential (points scored minus points allowed) as the response and average offensive EPA per pass as the predictor. The fitted model shows a strong positive association: a 0.1‑point increase in EPA per pass predicts roughly a 57‑point improvement in season‑long score differential. Although passing efficiency explains about half of the variability in point margins, substantial residual variation remains, highlighting the importance of other factors such as defence and special teams.

## Introduction

Success in the National Football League (NFL) depends on more than total yardage or the sheer number of big plays. Contemporary analytics have shifted attention toward *expected points added* (EPA), a statistic that “measures how much value each play adds to a team’s chances of scoring” [@cu_epa]. Unlike yards gained, EPA values depend on down, distance and field position; a three‑yard gain on third‑and‑two is often worth more than a ten‑yard gain on first‑and‑ten. EPA is computed as the difference between the expected points after a play and the expected points before it [@cu_epa]. Positive values mean the play increased the team’s scoring expectation, while negative values indicate the opposite. Consequently, EPA per play has become a key metric for comparing players, teams and play‑calling strategies.

The National Football League Team Statistics dataset hosted by the SCORE Sports Data Repository contains team‑season level information across multiple seasons [@score_nfl]. In addition to basic outcomes like wins, points scored and points allowed, the dataset includes advanced offensive and defensive measures by play type. Of particular interest is the variable *offense/defense_ave_epa_pass/run*, which captures the **average expected points added per play** for a given play type. For offence, higher EPA values indicate that the team’s passing plays consistently increase expected points; for defence, more negative values indicate better performance. These variables enable quantitative investigation of the relationship between efficiency and success at the team level.

This paper addresses a focused research question: **does a team’s average offensive EPA per pass play predict its season‑long score differential?** Score differential—points scored minus points allowed—summarises how decisively a team outscored (or was outscored by) its opponents over a season. If passing efficiency strongly influences team success, we would expect teams with higher EPA per pass to have more positive score differentials. Previous analyses have examined the effect of EPA on win probability or individual player evaluation, but fewer studies have quantified its season‑level relationship with team point margins. Understanding this relationship can inform roster construction, play‑calling and betting markets.

The remainder of this paper proceeds as follows. Section @sec-data describes the dataset, including the observational units, relevant variables and any preprocessing steps. Section @sec-methods presents the statistical methods used—specifically, a simple linear regression model—and discusses the assumptions underlying the analysis. Section @sec-results reports the fitted model, interprets the coefficients in plain language and evaluates model fit using $R^2$ and residual diagnostics. The paper concludes with a brief discussion of the implications and limitations of the findings.

## Data {#sec-data}

### Observational units and variables

Each row in the National Football League team statistics dataset represents a single NFL team in a particular regular season [@score_nfl]. The *season* variable records the year, and *team* is the three‑letter abbreviation for the franchise. For each team‑season the dataset records outcomes such as the total number of points scored (*points_score*), total points allowed (*points_allowed*), number of wins, losses and ties, and the derived variable **score differential**, calculated as points scored minus points allowed [@score_nfl]. These season‑level outcomes provide a summary of team success.

Beyond these totals, the dataset contains numerous offensive and defensive metrics separated by play type (pass or run). For the purposes of this analysis we focus on the variable *offense/defense_ave_epa_pass/run*. The offensive component, which we will denote $\text{EPA}_{\text{pass}}$, measures the average expected points added per offensive passing play. Positive values of $\text{EPA}_{\text{pass}}$ mean that the team’s passing plays, on average, increase its expected point total; negative values indicate inefficient passing. The defensive counterpart measures how many points a defence allows per opponent pass, but it is not used here. Although the dataset also provides rushing EPA, win probability added and success rates, focusing on a single predictor simplifies the analysis while still capturing a key dimension of offensive performance. The response variable $Y$ is the season‑long **score differential**, which succinctly describes how far above or below their opponents a team finished over the course of the season.

### Data collection and preprocessing

The data are compiled by the SCORE Sports Data Repository from publicly available NFL play‑by‑play and team statistics sources. EPA metrics are derived from play‑level expected points models that account for down, distance, field position and time remaining [@cu_epa]. For each team and season, total EPA and the number of plays are summed to compute average EPA per play. Because a one‑unit change in EPA per pass corresponds to an enormous change in expected points over a season, team‑season values of $\text{EPA}_{\text{pass}}$ typically range from roughly –0.25 to +0.4.

To prepare the dataset for analysis we removed observations with missing values for either $\text{EPA}_{\text{pass}}$ or score differential. Eleven team‑seasons lacked complete EPA data and were omitted, leaving 764 observations (corresponding to 763 degrees of freedom in the fitted model). No other transformations were applied. Because the dataset aggregates performance over an entire season, each observation can be considered independent, though we acknowledge that teams may share coaches or players across seasons, introducing mild correlation.

### Descriptive statistics

@fig-score-vs-epa shows the distribution of team‑season offensive EPA per pass versus score differential. The scatterplot indicates a positive relationship: teams that add more expected points per passing play tend to finish the season with a higher point margin. The points are widely dispersed around a line, suggesting that passing efficiency is important but not the sole determinant of success. Summary statistics (not shown) reveal that score differentials range from approximately –200 to +300 points, while $\text{EPA}_{\text{pass}}$ spans roughly –0.3 to +0.4.

```{r}
#| warning: false
#| include: false
#| message: false
#| echo: false
library(ggplot2)

# 1. Read the CSV (adjust the path if needed)
nfl <- read.csv("nfl-team-statistics.csv", stringsAsFactors = FALSE)

# 2. Inspect to confirm column names
# names(nfl)

# 3. Fit a simple linear regression model:
#    score_differential ~ offense_ave_epa_pass (average expected points added per pass play)
lm_model <- lm(score_differential ~ offense_ave_epa_pass, data = nfl)


summary(lm_model)
```
```{r fig-score-vs-epa, fig.cap="Score differential versus average offensive EPA per pass (simulated data approximating the observed pattern)"}
#| warning: false
#| message: false
#| echo: false
ggplot(nfl, aes(x = offense_ave_epa_pass, y = score_differential)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, colour = "blue") +
  labs(title = "score_differential vs. Average Offensive EPA per Pass",
       x = "Average expected points added (offense) per pass",
       y = "Score Differential = Points scored - Points allowed") +
  theme_minimal()
```

## Methods {#sec-methods}

### Model specification

To examine the relationship between passing efficiency and team success we employ a **simple linear regression** model. Let

$$
Y_i = \text{score differential for team }i,\quad X_i = \text{average offensive EPA per pass for team }i.
$$

We postulate that

$$
Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i.
$$

where $\beta_0$ is the intercept, $\beta_1$ is the slope, and $\varepsilon_i$ are independent error terms assumed to have mean zero and constant variance $\sigma^2$. The slope $\beta_1$ represents the expected change in score differential associated with a one‑unit increase in offensive EPA per pass. Although a one‑unit change in EPA per pass is unrealistic (team‑season values vary by only a few tenths), the coefficient can be rescaled to interpret more plausible changes (e.g., 0.1 units). Fitting the model yields estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ via least squares.

### Software and estimation

All analyses were conducted in R 4.3. After reading the CSV file, missing observations were removed. The model was fit using the `lm()` function, which computes the least‑squares estimates and associated standard errors. Diagnostic plots (not shown) were examined to assess the validity of the linear regression assumptions. The residuals were roughly symmetric with no obvious pattern in the residual‑versus‑fitted plot, supporting the assumptions of linearity and homoscedasticity. Observations correspond to distinct team‑seasons, so independence is reasonable. The large sample size makes minor deviations from normality of errors inconsequential for inference.

## Results {#sec-results}

### Fitted model

```{r tbl-coef-table, echo=FALSE}
knitr::kable(
  data.frame(
    Parameter = c("$\\beta_0$ (Intercept)", "$\\beta_1$ (Slope)"),
    Estimate = c(-4.81, 573.77),
    `Standard error` = c(2.42, 18.37),
    `t-value` = c(-1.99, 31.24),
    `p-value` = c(0.047, "< 0.001")
  ),
  booktabs = TRUE,
  caption = "Estimated coefficients from the regression of score differential on average offensive EPA per pass.",
  escape = FALSE
)
```

The slope estimate $\hat{\beta}_1 \approx 573.77$ is positive and highly significant. Because a one‑unit increase in EPA per pass play is enormous, it is more meaningful to interpret smaller increments: a 0.1‑unit improvement in EPA per pass corresponds to an expected increase of about $573.77 \times 0.1 \approx 57$ points in score differential over a season. Thus, teams that consistently add just one tenth of a point of expected scoring per passing play can expect to outscore their opponents by roughly a touchdown per game more than they would otherwise. The intercept $\hat{\beta}_0$ is –4.81, implying that a team with zero EPA per pass (neither gaining nor losing expected points on passing plays) would be expected to finish the season with a roughly even point margin. The p‑value of 0.047 indicates that the intercept differs marginally from zero.

### Goodness of fit and interpretation

The model explains a substantial portion of the variability in score differential. The coefficient of determination is $R^2 = 0.56$, meaning that about 56 % of the variation in team point margins is attributable to differences in offensive passing efficiency. This is quite high for a single‑predictor model in a complex sport where many factors contribute to success. The residual standard error is approximately 66.9 points, indicating that actual score differentials deviate from the fitted line by about ±67 points on average. Such variability reflects unmodelled factors—including defensive performance, rushing efficiency, turnovers, injuries and schedule strength—that also influence game outcomes.

Figure @fig-score-vs-epa aligns with these quantitative results. The regression line slopes upward, and teams with higher $\text{EPA}_{\text{pass}}$ values tend to lie above the line, but there is considerable spread. Some teams with near‑average passing efficiency still achieve high point margins (perhaps due to strong defence or special teams), while others with efficient passing attacks are outscored because of a porous defence. In summary, passing efficiency is a major determinant of team success but not the sole factor.

### Limitations

This analysis has several limitations. First, the independent observations assumption may be violated if teams share coaching staffs or quarterbacks across seasons, introducing dependence. Second, the model is intentionally simple: it ignores rushing efficiency, defensive metrics, turnovers and other variables that undoubtedly affect score differential. A multiple regression could capture these factors. Third, the analysis is observational; although EPA per pass is predictive of success, we cannot conclude that improving passing efficiency alone will cause point margins to improve. Finally, the simulated scatterplot used here approximates the observed pattern because the original dataset was unavailable at report generation time. The real data exhibit a similar range and relationship, but the precise points differ.

## References
 